Master Thesis

### 2013-06-17 acg.tex
This is the printed version.

Let us take a look at an example from
\cite{pogodalla2007generalizing}. We will start with a very simple
signature of string expressions with the concatenation operator,
$\Sigma_{String} = \mathopen{<}A_{String}, C_{String},
\tau_{String}\mathclose{>}$. Our signature will need only one atomic
type, $A_{String} = \{\textsc{String}\}$. Our constants will be the
wordforms of our example fragment (\emph{every}, \emph{some},
\emph{man}, \emph{woman}, \emph{loves}), the empty string $\epsilon$ and
the string concatenation operator, $+$. $\tau_{String}$ assigns the type
$\textsc{String}$ to all the wordforms and to $\epsilon$ and the type
$\textsc{String} \limp \textsc{String} \limp
\textsc{String}$\footnote{As a convention, whenever we omit parentheses
  in a type, we presume the $\limp$ and $\to$ type constructors always
  bind from the right, meaning that $a \to b \to c$ should be read as $a
  \to (b \to c)$.}  to the concatenation operator $+$.

@@

\begin{itemize}
\item $G$ is a mapping from $C_1$ to $\Lambda(\Sigma_2)$ assigning to
  every constant of the abstract signature a term in the object
  signature, which can be understood as its
  interpretation/implementation/realization.
\item $F$ is a mapping from $A_1$ to $\mathcal{T}(A_2)$ which links the
  abstract-level types with the object-level types that they realized
  as.
\item $F$ and $G$ are compatible, meaning that for any $c \in C_1$, we
  have $\vdash_{\Sigma_2} G(c) : \hat{F}(\tau_1(c))$ (we will be using
  $\hat{F}$ and $\hat{G}$ to refer to the homomorphic extensions of $F$
  and $G$ to $\mathcal{T}(A_1)$ and $\Lambda(\Sigma_1)$ respectively).
\end{itemize}

@@

Our lexicon $\mathcal{L}_{syntax}$ will map the constants of
$\Sigma_{Synt}$, our abstract signature, into terms from
$\Lambda(\Sigma_{String})$, our object signature. If we view terms of
$\Lambda(\Sigma_{Synt})$ as abstract computations, the lexicon will
instantiate this abstraction by providing an implementation for the
constants of $\Sigma_{Synt}$. This way we can map an abstract
computation representing a phrase into a more specific object
computation that computes the string representing the phrase.

@@

Given the above mappings, we can use the homomorphic properties of our
type mapping to arrive at the object-level interpretations of complex
abstract types, such as the type of predicates, $NP \limp S$, which
becomes a unary string function, $\textsc{String} \limp
\textsc{String}$, or the type of quantified noun phrases, $(NP \limp S)
\limp S$, which becomes a higher-order string function,
$(\textsc{String} \limp \textsc{String}) \limp \textsc{String}$.

>>>>

Let us take a look at an example from
\cite{pogodalla2007generalizing}. We will start with a very simple
signature of string expressions with the concatenation operator,
$\Sigma_{String} = \mathopen{<}A_{String}, C_{String},
\tau_{String}\mathclose{>}$. Our signature will need only one atomic
type, $A_{String} = \{\textsc{String}\}$. Our constants will be the
wordforms of our example fragment (\emph{every}, \emph{some},
\emph{man}, \emph{woman}, \emph{loves}), the empty string $\epsilon$ and
the string concatenation operator, $+$. $\tau_{String}$ assigns the type
$\textsc{String}$ to all the wordforms and to $\epsilon$, and the type
$\textsc{String} \limp \textsc{String} \limp
\textsc{String}$\footnote{As a convention, whenever we omit parentheses
  in a type, we presume the $\limp$ and $\to$ type constructors always
  bind from the right, meaning that $a \to b \to c$ should be read as $a
  \to (b \to c)$.}  to the concatenation operator $+$.

@@

\begin{itemize}
\item $G$ is a mapping from $C_1$ to $\Lambda(\Sigma_2)$ assigning to
  every constant of the abstract signature a term in the object
  signature, which can be understood as its
  interpretation/implementation/realization.
\item $F$ is a mapping from $A_1$ to $\mathcal{T}(A_2)$ which links the
  abstract-level types with the object-level types that they realized
  as.
\item $F$ and $G$ are compatible, meaning that for any $c \in C_1$, we
  have $\vdash_{\Sigma_2} G(c) : \hat{F}(\tau_1(c))$ (we will be using
  $\hat{F}$ and $\hat{G}$ to refer to the homomorphic extensions of $F$
  and $G$ to $\mathcal{T}(A_1)$ and $\Lambda(\Sigma_1)$
  respectively). This property will be referred to as the
  \emph{homomorphism property} of a lexicon.
\end{itemize}

@@

Our lexicon $\mathcal{L}_{syntax}$ will map the constants of
$\Sigma_{Synt}$, our abstract signature, into terms from
$\Lambda(\Sigma_{String})$, our object signature. If we view terms of
$\Lambda(\Sigma_{Synt})$ as abstract computations, the lexicon will
instantiate this abstraction by providing an implementation for the
constants of $\Sigma_{Synt}$. This way we can map an abstract
computation representing a phrase into a more specific object
computation which calculates the string representation.

@@

We can construct the homomorphic extension of the above type mapping
which can give us the object-level interpretations of complex abstract
types, such as the type of predicates, $NP \limp S$, which becomes a
unary string function, $\textsc{String} \limp \textsc{String}$, or the
type of quantified noun phrases, $(NP \limp S) \limp S$, which becomes a
higher-order string function, $(\textsc{String} \limp \textsc{String})
\limp \textsc{String}$.

### 2013-06-17 constraints.tex
This is the printed version.

TODO: Change this after splitting the chapters.

In the rest of this chapter, we will devote ourselves to investigating
ways of disentangling these.

\section{Intersections of Types}

Our goal is to split the types into smaller parts that only talk about
features belonging to one of the three mechanisms covered above. This
means that we would like to take some phrase about which we know a lot
of disparate pieces of information (presence of negative determiner,
trace status, number\ldots), having type
e.g. $NP\_\textcolor{red}{NEG{=}T}\_\textcolor{green}{VAR{=}F}\_\textcolor{blue}{NUM{=}PL}$,
and use it as an argument for some function which only cares about one
channel of information and would have a type like
$NP\_\textcolor{red}{NEG{=}T} \limp N\_\textcolor{red}{NEG{=}F} \limp
N\_\textcolor{red}{NEG{=}T}$. To make a type such as
$NP\_\textcolor{red}{NEG{=}T}\_\textcolor{green}{VAR{=}F}\_\textcolor{blue}{NUM{=}PL}$
be a legal argument to a function which expects an argument of type
$NP\_\textcolor{red}{NEG{=}T}$, we will need to introduce some notion of
\emph{subtyping}.

The types we have been showing up to now propose a very straightforward
notion of subtyping. All of the types can be seen as conjunctions of
properties where $\_$ (underscore) is the conjunction operator. A type
$\tau$ can then be seen as a subtype of type $\sigma$ whenever the
conjuncts of $\tau$ form a superset of the conjuncts of $\sigma$. This
is not unlike the structural typing of records in OCaml (CITE SOME OCAML
SPEC HERE), where a record type is a subtype of another record type if
the former provides, among others, all the fields provided by the latter
and gives them the same type. In our case, one type is a subtype of
another if the former gives, among others, all the properties given by
the latter. This notion of subtyping can be expressed by adding the rule
of inference in Figure~\ref{fig:type-superset} to our type
system.\footnote{In the rule \ldots (talk about $\land$ instead of $\_$
  (underscore) here...)}

\begin{figure}
  \begin{prooftree}
    \AxiomC{$\Gamma_l; \Gamma_i \vdash_\Sigma M : A_1 \land \ldots A_{i-1} \land A_i \land A_{i+1} \ldots \land A_n$}
    \RightLabel{(subtype)}
    \UnaryInfC{$\Gamma_l; \Gamma_i \vdash_\Sigma M : A_1 \land \ldots A_{i-1} \land A_{i+1} \ldots \land A_n$}
  \end{prooftree}
  \caption{\label{fig:type-superset} A rule of inference giving us
    subtyping for conjunctions of atomic types. For readability, we use
    $\land$ instead of $\_$ (underscore) to designate the conjunction of
    properties.}
\end{figure}

(Talk about whether or not this is a conservative extension...)

With the new subtyping rule, we can

$$
S_{tatou} : N \land NEG{=}F \land NUM{=}SG
$$
$$
S_{le_{NEG_1}} : (N \land NEG{=}F) \limp (((NP \land NEG{=}F) \limp S) \limp S)
$$
$$
S_{le_{NEG_2}} : (N \land NEG{=}T) \limp (((NP \land NEG{=}T) \limp S) \limp S)
$$
$$
S_{le_{NUM}} : (N \land NUM{=}SG) \limp (((NP \land NUM{=}SG) \limp S) \limp S)
$$
$$
S_{le_{VAR}} : N \limp (((NP \land VAR{=}F) \limp S) \limp S)
$$

\ldots

>>>>

In the next chapter, we will present our proposed solution to the
problem of writing simple grammars that cover multiple linguistic
constraints.

### 2013-06-17 gacg.tex
This is the printed version.

Further than that, we can look at the languages $\mathcal{L}_{12}(A_1)$
and $\mathcal{L}_{23}(A_2)$, images of the abstract languages given by
the lexicons. Since the lexicons are homomorphic mappings w.r.t. the
types, all elements of $\mathcal{L}_{12}(A_1)$ will have the type
$\mathcal{L}_{12}(S_1) = S_2$ and will therefore also belong to $A_2$
(similarly, we have that $\mathcal{L}_{23}(A_2) \subseteq A_3$). These
new languages, which are formed by mapping an ACG abstract language
using a lexicon, are called \emph{object languages}. Their terms are
constrained not only by the types of the signature they are built upon
but also by the type system of the abstract signature.

@@

We will associate a language to every node of an ACG diagram. Before we
begin, we notice that every ACG diagram formed by the above modes of
composition is an arborescence\footnote{An arborescence is a directed
  rooted tree, where every edge points away from the root.} and
therefore every node has at most one predecessor (the parent). To the
root of the tree, the abstract-most signature in the graph, we will
assign the abstract language generated by that signature and its
distinguished type. To all the other nodes, we will assign the result of
mapping their parent's language by the lexicon linking the two. By
homomorphism of the lexicons, we can find distinguished types for all
the non-root nodes. They are simply the images of their parents'
distinguished types by the connecting lexicons. For all of the nodes, it
is true that every term in the language assigned to the node has the
node's distinguished type.

@@

This definition works for non-root nodes of an ACG diagram. In the root
case, we have to constrain the language by stating that that the terms
of the language must have the root's distinguished type. This is a
property that is also true in the non-root cases due to the homomorphic
properties of lexicons. If we wanted to unify the definitions of the
root language and the non-root languages, we would have to check that a
term of the language has the distinguished type and that it has an
antecedent in the parent, if it has any parent. A relational statement
of this unification could look like this.

@@

  \begin{itemize}
    \item Let $L$ be an object language defined by the ACG $\mathcal{G}
      = \mathopen{<} \Sigma_A, \Sigma_O, \mathcal{L},
      S\mathclose{>}$. We take the graph $G$ with $V(G) = \{A, O\}$ and
      $E(G) = \{(A,O)\}$, we label the nodes $A$ and $O$ with $\Sigma_A$
      and $\Sigma_O$ as signatures and $S$ and $\mathcal{L}(S)$ as
      distinguished types respectively. Finally, we label the edge
      $(A,O)$ with $\mathcal{L}$ and we get a graphical ACG $G'$ in
      which $\mathcal{E}_{\mathcal{G'}}(O) = \mathcal{O}(G) = L$.

    \item Let $L_1$ and $L_2$ be two extrinsic languages defined by the
      nodes $u_1$ and $u_2$ in some graphical ACGs $\mathcal{G}_1$ and
      $\mathcal{G}_2$, respectively. We can construct a new graphical
      ACG $\mathcal{G}'$ by taking the union of the two ACGs (union of
      graphs (union of vertices and edges) and unions of the
      labelings).\footnote{Assuming, without loss of generality, that
        the sets of vertices of the two graphs are disjoint.} We will
      then add a new node $v$ such that $\mathcal{E}_{\mathcal{G}'}(v) =
      L_1 \cap L_2$.

@@

      \begin{align*}
        \mathcal{E}_{\mathcal{G}'}(v) &= \mathcal{I}_{\mathcal{G}'}(v)
        \cap \mathcal{L}_{(u_1,v)}(\mathcal{E}_{\mathcal{G}'}(u_1)) \cap
        \mathcal{L}_{(u_2,v)}(\mathcal{E}_{\mathcal{G}'}(u_2)) \\ &=
        \mathcal{I}_{\mathcal{G}'}(v) \cap
        \mathcal{E}_{\mathcal{G}'}(u_1) \cap
        \mathcal{E}_{\mathcal{G}'}(u_2) \\ &=
        \mathcal{E}_{\mathcal{G}'}(u_1) \cap
        \mathcal{E}_{\mathcal{G}'}(u_2) \\ &=
        \mathcal{E}_{\mathcal{G}}(u_1) \cap
        \mathcal{E}_{\mathcal{G}}(u_2) \\ &= L_1 \cap L_2
      \end{align*}

      The simplifications use the facts that: the two new lexicons are
      identities, that $\mathcal{I}_{\mathcal{G}'}(v) \supseteq
      \mathcal{E}_{\mathcal{G}'}(u_1) \cap
      \mathcal{E}_{\mathcal{G}'}(u_2)$, that
      $\mathcal{E}_{\mathcal{G}'}(u_1) = \mathcal{E}_{\mathcal{G}}(u_1)$
      and $\mathcal{E}_{\mathcal{G}'}(u_2) =
      \mathcal{E}_{\mathcal{G}}(u_2)$ and finally that
      $\mathcal{E}_{\mathcal{G}}(u_1) = L_1$ and
      $\mathcal{E}_{\mathcal{G}}(u_2) = L_2$.

  \item Let $L \subseteq \Lambda(\Sigma_O)$ be a language which is the
    result of mapping an extrinsic language $L' \subseteq
    \Lambda(\Sigma_A)$ through the lexicon $\mathcal{L} : \Sigma_A \to
    \Sigma_O$. We will show that $L$ is also an extrinsic language. Let
    $u$ be the node that defines the extrinsic language $L$ in some
    graphical ACG $\mathcal{G}$. We construct a new graphical ACG
    $\mathcal{G}'$ by adding a new vertex $v$ to $\mathcal{G}$ labeled
    with the signature $\Sigma_O$ and the distinguished type
    $\mathcal{L}(S)$ where $S$ is the distinguished type labeling
    $u$. To this graph, we conjoin the edge $(u,v)$ and label it with
    $\mathcal{L}$. Now we verify that $\mathcal{E}_{\mathcal{G}'}(v) =
    L$.

>>>>

Further than that, we can look at the languages $\mathcal{L}_{12}(A_1)$
and $\mathcal{L}_{23}(A_2)$, images of the abstract languages given by
the lexicons. Because of the homomorphism property of lexicons, all
elements of $\mathcal{L}_{12}(A_1)$ will have the type
$\mathcal{L}_{12}(S_1) = S_2$ and will therefore also belong to $A_2$
(similarly, we have that $\mathcal{L}_{23}(A_2) \subseteq A_3$). These
new languages, which are formed by mapping an ACG abstract language
using a lexicon, are called \emph{object languages}. Their terms are
constrained not only by the types of the signature they are built upon
but also by the type system of the abstract signature.

@@

We will associate a language to every node of an ACG diagram. Before we
begin, we notice that every ACG diagram formed by the above modes of
composition is an arborescence\footnote{An arborescence is a directed
  rooted tree, where every edge points away from the root.} and
therefore every node has at most one predecessor (the parent). To the
root of the tree, the abstract-most signature in the graph, we will
assign the abstract language generated by that signature and its
distinguished type. To all the other nodes, we will assign the result of
mapping their parent's language by the lexicon linking the two. Thanks
to the homomorphism property of the lexicons, we can find distinguished
types for all the non-root nodes. They are simply the images of their
parents' distinguished types by the connecting lexicons. For all of the
nodes, it is true that every term in the language assigned to the node
has the node's distinguished type.

@@

This definition works for non-root nodes of an ACG diagram. In the root
case, we have to constrain the language by stating that that the terms
of the language must have the root's distinguished type. This is a
property that is also true in the non-root cases due to the homomorphism
property of lexicons. If we wanted to unify the definitions of the root
language and the non-root languages, we would have to check that a term
of the language has the distinguished type and that it has an antecedent
in the parent, if it has any parent. A relational statement of this
unification could look like this.

@@

  \begin{figure}
    \centering
    \begin{subfigure}[b]{0.2\textwidth}
      \centering
      \includegraphics[height=0.3\textheight]{diagrams/proof-object.pdf}
      \caption{\label{fig:proof-object} The case for object languages.}
    \end{subfigure}
    \quad
    \begin{subfigure}[b]{0.4\textwidth}
      \centering
      \includegraphics[height=0.3\textheight]{diagrams/proof-intersection.pdf}
      \caption{\label{fig:proof-intersection} The case for intersections.}
    \end{subfigure}
    \quad
    \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[height=0.3\textheight]{diagrams/proof-lexicon.pdf}
      \caption{\label{fig:proof-lexicon} The case for transformations by
        a lexicon.}
    \end{subfigure}
    \caption{Figures demonstrating that extrinsic languages contain
      object languages closed on transformation by a lexicon and
      intersection.}
  \end{figure}

  \begin{itemize}
    \item Let $L$ be an object language defined by the ACG $\mathcal{G}
      = \mathopen{<} \Sigma_A, \Sigma_O, \mathcal{L},
      S\mathclose{>}$. We take the graph $G$ with $V(G) = \{A, O\}$ and
      $E(G) = \{(A,O)\}$, we label the nodes $A$ and $O$ with $\Sigma_A$
      and $\Sigma_O$ as signatures and $S$ and $\mathcal{L}(S)$ as
      distinguished types respectively . Finally, we label the edge
      $(A,O)$ with $\mathcal{L}$ and we get a graphical ACG $G'$ in
      which $\mathcal{E}_{\mathcal{G'}}(O) = \mathcal{O}(G) = L$
      (Figure~\ref{fig:proof-object}).

    \item Let $L_1$ and $L_2$ be two extrinsic languages defined by the
      nodes $u_1$ and $u_2$ in some graphical ACGs $\mathcal{G}_1$ and
      $\mathcal{G}_2$, respectively. We can construct a new graphical
      ACG $\mathcal{G}'$ by taking the union of the two ACGs (union of
      graphs (union of vertices and edges) and unions of the
      labelings).\footnote{Assuming, without loss of generality, that
        the sets of vertices of the two graphs are disjoint.} We will
      then add a new node $v$ such that $\mathcal{E}_{\mathcal{G}'}(v) =
      L_1 \cap L_2$ (Figure~\ref{fig:proof-intersection}).

@@

      \begin{align*}
        \mathcal{E}_{\mathcal{G}'}(v) &= \mathcal{I}_{\mathcal{G}'}(v)
        \cap \mathcal{L}_{(u_1,v)}(\mathcal{E}_{\mathcal{G}'}(u_1)) \cap
        \mathcal{L}_{(u_2,v)}(\mathcal{E}_{\mathcal{G}'}(u_2)) \\ &=
        \mathcal{I}_{\mathcal{G}'}(v) \cap
        \mathcal{E}_{\mathcal{G}'}(u_1) \cap
        \mathcal{E}_{\mathcal{G}'}(u_2) \\ &=
        \mathcal{E}_{\mathcal{G}'}(u_1) \cap
        \mathcal{E}_{\mathcal{G}'}(u_2) \\ &=
        \mathcal{E}_{\mathcal{G}_1}(u_1) \cap
        \mathcal{E}_{\mathcal{G}_2}(u_2) \\ &= L_1 \cap L_2
      \end{align*}

      The simplifications use the facts that: the two new lexicons are
      identities, that $\mathcal{I}_{\mathcal{G}'}(v) \supseteq
      \mathcal{E}_{\mathcal{G}'}(u_1) \cap
      \mathcal{E}_{\mathcal{G}'}(u_2)$, that
      $\mathcal{E}_{\mathcal{G}'}(u_1) =
      \mathcal{E}_{\mathcal{G}_1}(u_1)$ and
      $\mathcal{E}_{\mathcal{G}'}(u_2) =
      \mathcal{E}_{\mathcal{G}_2}(u_2)$ and finally that
      $\mathcal{E}_{\mathcal{G}}(u_1) = L_1$ and
      $\mathcal{E}_{\mathcal{G}}(u_2) = L_2$.

  \item Let $L \subseteq \Lambda(\Sigma_O)$ be a language which is the
    result of mapping an extrinsic language $L' \subseteq
    \Lambda(\Sigma_A)$ through the lexicon $\mathcal{L} : \Sigma_A \to
    \Sigma_O$. We will show that $L$ is also an extrinsic language. Let
    $u$ be the node that defines the extrinsic language $L$ in some
    graphical ACG $\mathcal{G}$. We construct a new graphical ACG
    $\mathcal{G}'$ by adding a new vertex $v$ to $\mathcal{G}$ labeled
    with the signature $\Sigma_O$ and the distinguished type
    $\mathcal{L}(S)$ where $S$ is the distinguished type labeling
    $u$. To this graph, we conjoin the edge $(u,v)$ and label it with
    $\mathcal{L}$ (Figure~\ref{fig:proof-lexicon}).

    Now we verify that $\mathcal{E}_{\mathcal{G}'}(v) = L$.

### 2013-06-17 ig.tex
This is the printed version.

We will briefly discuss one more grammar formalism and that is the
formalism of Interaction Grammars \cite{guillaume2009interaction}.
Interaction grammars are a grammatical formalism centered around the
concept of \emph{polarity}. An interaction grammar is a set of
(under-specified) tree descriptions which define a set of trees that can
be constructed by superposing some of these tree descriptions while
respecting the polarities described in the nodes of the individual tree
descriptions.

>>>>

We will briefly discuss one more grammar formalism and that is the
formalism of interaction grammars \cite{guillaume2009interaction}. This
formalism is of particular interest to us since it is based on the same
logical basis as ACGs and there is already a detailed wide-coverage
grammar written in it.

Interaction grammars are a grammatical formalism centered around the
concept of \emph{polarity}. An interaction grammar is a set of
(under-specified) tree descriptions which define a set of trees that can
be constructed by superposing some of these tree descriptions while
respecting the polarities described in the nodes of the individual tree
descriptions.

### 2013-06-17 implementation.tex
This is the printed version.

The third system was XMG, a language for writing metagrammars, which is
another goal of our system, the one which distinguishes it from the
original ACG toolkit. We believe that using a declarative programming
system such as miniKanren with the ability to extend it using a
practical functional programming language presents an empowering way of
defining metagrammars. This allows the grammar designer to compose his
grammar in any way he sees fit and to use the metaprogramming facilities
of the underlying programming language to transcribe the grammar in a
way that makes the most sense in her case.

@@

\begin{description}
  \item[Function of wordform] We could imagine that every different
    wordform of the lexical database will end up being mapped to some
    constant in the signature. However, this is unsuitable since the
    lexical database is a relation that can assign more than one
    distinct hypertag to a single wordform (e.g. a wordform with
    different possible part of speech) and we would like to have
    constants of different types for each of these hypertags.
  \item[Function of hypertag] Conversely, we might imagine that the
    hypertags of the lexical database are mapped via some function to
    the constants of the signature. Here we run into trouble when we
    consider variations in spelling or phonology. We might have two
    items in our lexical database that share the same hypertag but whose
    wordforms still differ. We would then like to have two distinct
    constants in our signature as well, so that when these two constants
    are mapped by the surface lexicon, they will yield two different
    strings.
  \item[Function of lexical entry] Since neither the wordform nor the
    hypertag is enough, we could consider the entire lexical entry (an
    element of the lexical database, a pair of a wordform and one of its
    hypertags). This solves both of the above problems but it is still
    not perfect. Consider the case of a determiner and some semantic
    signature. Since our link between elements of the lexical database
    and of our signature is a function, there must be a constant in the
    signature for every item of our lexical database. If you recall
    in~\ref{ssec:acg-sem}, the constants of the semantic signature
    $\Sigma_{Sem}$ contained predicates for the common nouns \emph{man}
    and \emph{woman} and for the verb \emph{loves} and some common
    logical furniture. However, it did not contain any constants which
    would correspond to the determiners \emph{every} and \emph{some} and
    therefore there would be no constant that the function could assign
    to these two wordforms.
  \item[Partial function of lexical entry] We could fix the issue above
    by using a partial function. This accounts for cases when a lexical
    entry has no constants to represent it in a signature. Nevertheless,
    this only fixes half of the problem. We will consider the single
    lexical entry which represents e.g. the 3rd person singular present
    tense form \emph{loves} of the transitive verb \emph{love}. A
    legitimate account of subject/object scope ambiguity would be to
    consider two constants in our abstract syntactic signature that have
    distinct images in some semantic lexicon, one giving the outer scope
    to the subject and the other to the object. This means that in order
    to define such a signature, we would need to link a single lexical
    entry in our database to more than one constant in the signature.
  \item[Relation between lexical entry and constant] Finally,
    considering all the above scenarios, the only mathematical model
    general enough to handle all of these cases is a relation. In our
    system, this will correspond exactly to a specific relation that
    will be provided by the grammar author.
\end{description}

@@

We have established that signatures will be defined by a relation, a
relation between a lexical entry (a wordform and a hypertag) and a
constant (an identity\footnote{The identity of a constant is simply the
  thing that distinguishes it from other constants of the same signature
  having the same type.} and a type). The identity of a lexical constant
also has a complex structure. We know that the lexical constants we
define using the relation described above are always linked to a lexical
entry. This lexical entry at least partially identifies the
constant. For cases where we have more than one constant per lexical
entry, such as in the example of the two different scope readings of
\emph{loves}, we need some extra piece of information to specify exactly
which of the two constants we are referring to (we call this extra piece
of information a \emph{specifier}). This specifier can have any shape
that the grammar author desires. We can see an example of a lexical
constant in Figure~\ref{fig:lex-const}.

@@

The system takes care of wiring up the relations and assembling the
signature. The grammar author's duty is to provide a 4-ary relation that
links the four parts of a lexical constant: the wordform, the hypertag,
the specifier and the type. This relation can be thought of as a
multi-valued ``function'' that receives as input a wordform and a
hypertag (that is, a single lexical entry) and outputs (possibly
multiple) pairs of specifiers and types.

Since we are working in a functional programming language that supports
manipulation of higher-order functions, we can provide helper functions
which construct these 4-ary relations for the grammar author in some
simple cases. Our library thus provides functions like
\texttt{unitypedr}, which expects a type and returns the 4-ary relation
that assigns to every lexical item a single constant whose specifier is
$nil$ and whose type is the supplied type, or \texttt{ht->typer}, which
expects a mapping from hypertag patterns to types and returns the 4-ary
relation which tries to map the hypertag against the patterns and yields
constants with the corresponding types and $nil$ specifiers. When other
patterns emerge, the grammar author is completely free to implement her
new abstractions using the full power of the functional programming
language.

@@

\subsection{Defining Lexicons}

@@

Our representation of lexical constants was purposefully engineered so
as to make the lexical entry directly accessible to the lexicon
implementation. Since the lexicon is directly implemented by the grammar
author as a binary relation between constants and terms, the grammar
author can inspect the abstract constant and directly access the
hypertag and wordform of the lexical entry that generated it, the
specifier that was assigned to it and the type of the constant and then
express the object term assigned to the constant using any combination
of these.

@@

The above is already enough to define any lexicon. However, our toolkit
presents convenience facilities that capture common forms of
lexicons. With these, we will be able to define lexicalized versions of
the lexicons of section~\ref{sec:acg} in a concise point-free
style. Furthermore, the grammar author is not limited to the convenience
facilities we have provided and is completely free to use the same
programming language we have used to build on our abstractions and build
her own right within her grammar definition.

@@

\section{Testing Signatures and Lexicons}

>>>>

The third system was XMG, a language for writing metagrammars, which is
another goal of our system, the one which distinguishes it from the
original ACG toolkit. We believe that using a declarative programming
system such as miniKanren with the ability to extend it using a
practical functional programming language presents an empowering way of
defining metagrammars. This allows the grammar designer to compose her
grammar in any way she sees fit and to use the metaprogramming
facilities of the underlying programming language to transcribe the
grammar in a way that makes the most sense in her case.

@@

\begin{description}
  \item[Function of wordform] We could imagine that every different
    wordform of the lexical database will end up being mapped to some
    constant in the signature. However, this is unsuitable since the
    lexical database is a relation that can assign more than one
    distinct hypertag to a single wordform (e.g. a wordform with
    different possible part of speech) and we would like to have
    constants of different types for each of these hypertags.
  \item[Function of hypertag] Conversely, we might imagine that the
    hypertags of the lexical database are mapped via some function to
    the constants of the signature. Here we run into trouble when we
    consider variations in spelling or phonology. We might have two
    items in our lexical database that share the same hypertag but whose
    wordforms still differ. We would then like to have two distinct
    constants in our signature as well, so that when these two constants
    are mapped by the surface lexicon, they will yield two different
    strings.
  \item[Function of lexical entry] Since neither the wordform nor the
    hypertag is enough, we could consider the entire lexical entry (an
    element of the lexical database, a pair of a wordform and one of its
    hypertags). This solves both of the above problems but it is still
    not perfect. Consider the case of a determiner and some semantic
    signature. Since our link between elements of the lexical database
    and of our signature is a function, there must be a constant in the
    signature for every item of our lexical database. If you recall
    in~\ref{ssec:acg-sem}, the constants of the semantic signature
    $\Sigma_{Sem}$ contained predicates for the common nouns \emph{man}
    and \emph{woman} and for the verb \emph{loves} and some common
    logical furniture. However, it did not contain any constants which
    would correspond to the determiners \emph{every} and \emph{some} and
    therefore there would be no constant that the function could assign
    to these two wordforms.
  \item[Partial function of lexical entry] We could fix the issue above
    by using a partial function. This accounts for cases when a lexical
    entry has no constants to represent it in a signature. Nevertheless,
    this only fixes half of the problem. We will consider the single
    lexical entry which represents e.g. the 3rd person singular present
    tense form \emph{loves} of the transitive verb \emph{love}. A
    legitimate account of subject/object scope ambiguity would be to
    consider two constants in our abstract syntactic signature that have
    distinct images in some semantic lexicon, one giving the outer scope
    to the subject and the other to the object. This means that in order
    to define such a signature, we would need to link a single lexical
    entry in our database to more than one constant in the signature.
  \item[Relation between lexical entry and constant] Finally,
    considering all the above scenarios, the only mathematical model
    general enough to handle all of these cases is a relation. In our
    system, this will correspond exactly to a specific relation that
    will be provided by the grammar designer.
\end{description}

@@

We have established that signatures will be defined by a relation, a
relation between a lexical entry (a wordform and a hypertag) and a
constant (an identity\footnote{The identity of a constant is simply the
  thing that distinguishes it from other constants of the same signature
  having the same type.} and a type). The identity of a lexical constant
also has a complex structure. We know that the lexical constants we
define using the relation described above are always linked to a lexical
entry. This lexical entry at least partially identifies the
constant. For cases where we have more than one constant per lexical
entry, such as in the example of the two different scope readings of
\emph{loves}, we need some extra piece of information to specify exactly
which of the two constants we are referring to (we call this extra piece
of information a \emph{specifier}). This specifier can have any shape
that the grammar designer desires. We can see an example of a lexical
constant in Figure~\ref{fig:lex-const}.

@@

The system takes care of wiring up the relations and assembling the
signature. The grammar designer's duty is to provide a 4-ary relation
that links the four parts of a lexical constant: the wordform, the
hypertag, the specifier and the type. This relation can be thought of as
a multi-valued ``function'' that receives as input a wordform and a
hypertag (that is, a single lexical entry) and outputs (possibly
multiple) pairs of specifiers and types.

Since we are working in a functional programming language that supports
manipulation of higher-order functions, we can provide helper functions
which construct these 4-ary relations for the grammar designer in some
simple cases. Our library thus provides functions like
\texttt{unitypedr}, which expects a type and returns the 4-ary relation
that assigns to every lexical item a single constant whose specifier is
$nil$ and whose type is the supplied type, or \texttt{ht->typer}, which
expects a mapping from hypertag patterns to types and returns the 4-ary
relation which tries to map the hypertag against the patterns and yields
constants with the corresponding types and $nil$ specifiers. When other
patterns emerge, the grammar designer is completely free to implement
her new abstractions using the full power of the functional programming
language.

@@

\subsection{Defining Lexicons}
\label{ssec:def-lex}

@@

Our representation of lexical constants was purposefully engineered so
as to make the lexical entry directly accessible to the lexicon
implementation. Since the lexicon is directly implemented by the grammar
designer as a binary relation between constants and terms, the grammar
designer can inspect the abstract constant and directly access the
hypertag and wordform of the lexical entry that generated it, the
specifier that was assigned to it and the type of the constant and then
express the object term assigned to the constant using any combination
of these.

@@

The above is already enough to define any lexicon. However, our toolkit
presents convenience facilities that capture common forms of
lexicons. With these, we will be able to define lexicalized versions of
the lexicons of section~\ref{sec:acg} in a concise point-free
style. Furthermore, the grammar designer is not limited to the convenience
facilities we have provided and is completely free to use the same
programming language we have used to build on our abstractions and build
her own right within her grammar definition.

@@

\section{Checking Signatures and Lexicons}

Since we let the grammar designer define signatures and lexicons using
arbitrary relations, she can also end up defining relations which cannot
be interpreted as well-formed signatures or lexicons. For this reason,
our toolkit provides a set of testing facilities which help check that
the signatures and lexicons are well-defined.

\subsection{Checking Properties of Large Structures}

As opposed to the existing \textbf{ACG development toolkit}, we cannot
easily enumerate all of the elements in a given signature or lexicon and
check that they are all consistent. The size of the lexicalized grammar
is too large to make this approach practical.

We solve this problem by demanding that the grammar designer names some
subset of the structure and testing is done only on that subset. In our
case, the most practical solution has been to let the designer just list
wordforms that generate a sufficient subset of the structure, i.e. the
designer lists one wordform per every category she has handled in her
grammar.

A more comfortable solution, which would also be more technically
involved, would be to have the system test the grammar on an
automatically deduced selection of constants such that they are
guaranteed to cover all the cases and code paths in the grammar
designer's definitions. Implementing such a system could still be
manageable given the small number of primitives the relational
programming system is built upon.

\subsection{Checking Signatures}
\label{ssec:check-sig}

The crucial property that we need to check for in signatures is that the
relation between constant identities and their types is a function. This
simply reduces to querying the system for the number of typed constants
belonging to the signature and having some specific identity. If this
number is one for all of our test cases, we have verified that (the
tested subset of) the signature assigns exactly one type to every one of
its constants.

\subsection{Checking Lexicons}

For lexicons, we check the three following properties:

\begin{description}
\item[The lexicon is a function,] assigning exactly one object term to
  each abstract constant. The verification proceeds analogously to the
  case of checking signatures (subsection \ref{ssec:check-sig}).

\item[The lexicon assigns only well-typed object terms.] This is a
  consequence of the homomorphism property of lexicons. We verify it by
  checking whether we can infer some type for every object term assigned
  to a constant in the test set.

\item[The lexicon has the homomorphism property.] We use the process
  that we described in subsection~\ref{ssec:def-lex} for inferring the
  type mapping of a lexicon from its term mapping.

  We first enumerate the pairs of abstract constants belonging to the
  test set and the object terms assigned to them by the lexicon. We then
  take the types of these pairs. Finally, we state, by using
  unification, that this set of pairs is a subset of the homomorphic
  extension of some mapping of atomic abstract types and let the
  relational programming system find us that mapping. If it succeeds, we
  have verified the homomorphism property of the lexicon.
\end{description}

The above technique of verifying the signature and lexicon definitions
by automated tests is not only useful for checking their
well-formedness. The grammar designer can go further and implement new
tests which demonstrate grammar-specific properties, an example of which
would be the $cat$-$funct$ principle in Frigram that imposes constraints
on the relationship between the $cat$ and $funct$ features in the EPTDs
of the grammar.
